Popularity Adjusted Block Models are Generalized Random Dot Product Graphs

### 2022-02-08 referee-response.Rmd

1. p4. What is a hollow graph?  
\textcolor{blue}{
We changed the terminology a bit here and throughout the paper and
specify that the analysis assumes $G$ is an unweighted and undirected
graph with no self-loops 
(no edge directly from itself to itself), 
so $A$ is binary and symmetric with zeros on the diagonal. The phrase
``hollow graph'' is no longer used.  }

2. p4. $G$ as a graph is non-oriented or undirected and $A$ is symmetric.  
\textcolor{blue}{
We have made minor edits clarifying the structure of $G$ and $A$. In
particular we assume throughout the paper that $G$ is undirected and $A$ is
symmetric. 
}

3. p4. Please define the set $[n]$.  
\textcolor{blue}{
We added $[n] = \{1, 2, ..., n\}$ at the beginning of Section~2.1. 
}

4. p4. Definition 1. Do you consider a probability distribution on
   $z_1, ..., z_n$? 
   Please give the range of variation of $(i, j)$ when you define $P_{ij}$.  
\textcolor{blue}{
The range of $(i, j)$ is now included in the definition; in particular
we wrote in Section 2.1 that ``$1 \leq i < j \leq n$.''
In regard to the probability distribution on the labels $z_1, ...,
z_n$, none of our theoretical results assume anything about the distribution
of the $\{z_i\}$; we do, however, occasionally assume that the $n_k$
are sufficiently large for each $k = 1,2,\dots, K$. More specifically 
the convergence rate of Theorem 5 depends on the $n_k$, i.e., for
convergence we require that $n_k \to \infty$ as $n \rightarrow
\infty$. In practice, we also require that each $n_k$ is large enough
so that the latent vectors in the $k$th community span its
subspace; this is a rather mild requirement, e.g., if the latent vectors in the $k$th community lie in 
general positions then we only require $n_k
\geq K$.}

@@

6. p5. Definition 2. I do not understand the definition of the subset $\mathcal{X}$. 
Is it $x \in \mathcal{X}$ if for all $y \in \mathbb{R}^d$ we have $x^\top I_{p,q} y$ or is $\mathcal{X}$ a subset on $(\mathbb{R}^d)^2$? 
Please clearly state that $d = p + q$ (not stated before Definition 3).  
\textcolor{blue}{
We now denote $d = p + q$ in Definition 2 and define $\mathcal{X}$ as a subset of $\mathbb{R}^{d}$ such
that for any $x \in \mathcal{X}$ and $y \in \mathcal{X}$, we have
$x^{\top} I_{p,q} y \in [0,1]$. One simple way to construct such a set
$\mathcal{X}$ is to first find a collection of $K$ vectors
$\mathcal{S} = \{\nu_1, \nu_2, \dots, \nu_K\} \subset \mathbb{R}^{d}$
where $\nu_i^{\top} I_{p,q} \nu_j \in [0,1]$ for all $i,j$ and then define $\mathcal{X}$
as 
$$\mathcal{X} = \mathrm{conv}(\mathcal{S}) = \{x = \sum_{i} \lambda_i \nu_i \colon \lambda_i \geq 0 \,\, \text{for all $i$}, \sum_{i} \lambda_i =
1\}.$$
}

7. p6. Remark 2. What are $\hat{Z}$, $V$, and $D$?  
\textcolor{blue}{
We had rewrote Remark~2 to more clearly described the matrices
$\hat{Z}$, $V$ and $D$. To avoid confusion, we now use $\hat{D}$ to
denote the diagonal matrix whose diagonal entries are the $d = p + q$
largest eigenvalues (in modulus) of $A$ and we use $\hat{V}$ to denote
the $n \times d$ matrix whose columns are the corresponding
eigenvectors. We then use $\hat{Z} = \hat{V} |\hat{D}|^{1/2}$ (where
the $|\cdot|$ operation is applied elementwise) to denote the
adjacency spectral embedding (ASE) of $A$ into $\mathbb{R}^{d}$; that is
to say, the $i$th row of $\hat{Z}$ represents an estimate of the
latent position $X_i$ (up to some non-identifiability transformation
$Q \in \mathbb{O}(p,q)$). We also reorganized the paper slightly so
that the definition and remarks about the non-identifiability of the
latent positions $X$ in a GRDPG are presented before discussing ASE. }

8. top of p7. $\tilde{P}$ instead of ${P}$.  
\textcolor{blue}{
We have corrected the typo. 
}

9. p7 line 40-46. I don't understand the interest of this paragraph.  
\textcolor{blue}{
The purpose of the additional material in the proof of Theorem 1 is to
first present the proof in the special case when $K = 2$ so as to
build intuition for the general case of $K \geq 2$. 
More specifically for $K = 2$ we show that $\tilde{P} = X \Pi X^\top$
where $\Pi$ is a permutation matrix with 2 fixed points and 1 cycle of
order 2, and the latent vectors lie in the union of two $2$-dimensional
orthogonal subspaces, i.e., the rows of $X$ consist of
vectors that lie in $\mathcal{S}_1 \cup \mathcal{S}_2$ where
$\mathcal{S}_1$ and $\mathcal{S}_2$ are both $2$-dimensional
subspaces of $\mathbb{R}^{4}$ with $\mathcal{S}_1 \cap \mathcal{S}_2 = \{0\}$ and $x^{\top} y
= 0$ for all $x \in \mathcal{S}_1, y \in \mathcal{S}_2$. 
This then generalizes to larger $K$ such that the rows of $X$ are now
vectors that lie in the union of $K$ orthogonal subspaces
$\mathcal{S}_1 \cup \mathcal{S}_2 \cup \dots \mathcal{S}_K$ where each
$\mathcal{S}_k$ is a $K$-dimensional subspace of $\mathbb{R}^{K^2}$,
and $\Pi$ is a permutation matrix with 
$K$ fixed points and $K (K - 1)$ cycles of order 2. The permutation
matrix $\Pi$ then has $K(K-1)/2$ eigenvalues equal to $-1$ (due
to the $K(K-1)$ cycles of order $2$) and $K(K+1)/2$ eigenvalues equal
to $1$ (due to the $K(K-1)$ cycles of order $2$ together with the $K$
fixed points). We can thus write $\Pi = U I_{K(K+1)/2,K(K-1)/2}
U^{\top}$. We have revised the paper to more clearly present the above
observations. See also Example~1 on page 9 of the revised manuscript
and the response to comment $10$ below.
}

10. p8, line 14-21. I don't see why the permutation given by $\Pi$ has $K$ fixed points.  
\textcolor{blue}{
Example 1 illustrates these fixed points explicitly for $K = 3$. 
More specifically, comparing the matrices $X$ and $Y$, we see that
they have three columns in common: the first, fifth, and ninth. 
The other columns are permuted. The proof of Theorem 1 then
generalizes this to arbitrary $K$. In particular by the construction
of $X$, each column of $X$ has exactly one non-zero block
corresponding to some vector $\lambda_{k \ell} \in \mathbb{R}^{n_k}$. The fixed
points of $\Pi$ are then the columns with indices $r(K+1) +1 $ for $0 \leq r
\leq K-1$; these columns contain the vectors $\lambda_{kk}$ for
$1 \leq k \leq K$. The cycles of order $2$ swap the columns of $X$
containing the vectors $\lambda_{k \ell}$ and $\lambda_{\ell k}$ for
the $\tbinom{K}{2}$ pairs $\{k,\ell\}$ with $k \not = \ell$. 
}

11. Section 3.1. I found it confusing that community detection algorithms are detailed when $P$ is assumed to be known and then methods for estimating $P$ are recalled. A sentence in the beginning of this section to expose what is presented could help the reader.  
\textcolor{blue}{
We thank the referee for pointing out this issue. We had revised the
manuscript to include the line "In our methods, the data that are
observed is only the adjacency matrix $A \sim
\mathrm{PABM}(\{\lambda^{(k \ell)}\}_K, \rho_n)$ along with an assumed
number of communities, $K$. To motivate our methods, we first
consider community detection and parameter estimation in the case
where we know the edge probability matrix $P$ beforehand, noting
that community memberships and popularity parameters are not
immediately discernible from $P$ itself. After establishing methods
for community detection and parameter estimation from $P$, we use the
consistency property of ASE to demonstrate that the same methods work
for $A$ almost surely as $n \to \infty$." Indeed, the assumption of
known $P$ helps motivate Algorithm $1$ through Algorithm~3 and the
theoretical results in Theorem~2. Theorem 3 and Theorem 4 then follows
by leveraging the consistency properties for adjacency spectral embedding
to show that community recovery and parameters estimation using $A$ is asymptotically equivalent to
community recover and paraemters estimation using the true but unknown $P$. 
}

@@

12. p10, line 12. $c^{(i)}$ is not defined.  
\textcolor{blue}{
$c^{(i)}$ is the $i^{th}$ entry of column vector $c$. 
This has been added to the paper.
}

13. p10. I didn't understand the exposition of the SSC algorithm.  
\textcolor{blue}{
We have added more details to our description of SSC to
provide further intuition/motivation behind the
algorithm. The description of SSC now reads:
``The SSC algorithm can be described as follows: 
Given \(X \in \mathbb{R}^{n \times d}\) with vectors
\(x_i^\top \in \mathbb{R}^d\) as rows of \(X\), the optimization problem
\(c_i = \arg\min_{c} \|c\|_1\) subject to \(x_i = X^\top c\) and
\(c^{(i)} = 0\), where $c^{(i)}$ is the $i^{th}$ entry of $c$, is solved for each \(i \in [n]\). 
The solutions are collected into matrix
\(C = \bigl[ c_1 \mid \cdots \mid c_n \bigr]^\top\) to
construct an affinity matrix \(B = |C| + |C^\top|\). If each \(x_i\) lies
exactly on one of \(K\) subspaces then \(B\) describes an undirected graph
consisting of {\em at least} \(K\) disjoint subgraphs, i.e., \(B_{ij} = 0\) if \(x_i, x_j\) lie on different subspaces. 
The intuition here is that vectors that lie on the same subspace can
be described as linear combinations of each other, assuming the number
of vectors in the subspace is greater than the dimensionality of the
subspace. Thus, for each $c_i$, $c_i^{(j)}$ is zero whenever $x_i$ and $x_j$ belong to different subspaces and may be nonzero otherwise. If \(X\) instead represents points near \(K\) subspaces with some noise, 
then this property will only hold approximately and 
a final graph partitioning step may be required 
(e.g., edge thresholding or spectral clustering).''
}

14. p11, line 31. "... SSC on the spectral embedding of $A$". Could you support this assertion by references?  
\textcolor{blue}{
There are no references for this, as it is, to the best of our knowledge, a novel idea. 
We emphasize that while SSC has been proposed for community detection
for PABM (see e.g., \citep{noroozi2019estimation}), these work apply SSC
directly on the rows of $A$ (which are binary vectors in
$\mathbb{R}^{n}$) rather than on the ASE of $A$ (which are real-valued
vectors in $\mathbb{R}^{d}$ where $d \ll n$). The motivation behind
our use of SSC comes  from Theorem 1 wherein we show that the latent
vectors for PABM lie on the union of orthogonal subspaces in $\mathbb{R}^{d}$ and hence,
by the consistency properties of ASE, the rows of the estimated
eigenvectors $\hat{V}$ also lies close to the union of orthogonal
subspaces in $\mathbb{R}^{d}$. In contrast, while the rows of $P$ do lie exactly on the
union of $K$ subspaces in $\mathbb{R}^{n}$, the rows of $A$
(which are now noisy binary vectors) could be quite far from the
corresponding rows of $P$ and hence need not lie close to the union
of any $K$ subspaces in $\mathbb{R}^{n}$ 
}

@@

15. p12, line 13. It could help recall what the $z_i, z_j$ are.  
\textcolor{blue}{
We have changed $z_i$ to ``community labels''.
}

16. p14. Do you think that the OSC algorithm can retrieve other kinds of structure than assortative ones?  
\textcolor{blue}{
The OSC algorithm works for any PABM parameters setting and is thus
agnostic to the particular structure (such as assortativity vs
disassortativity). We had revised the manuscript to include an
additional simulation demonstrating this claim (see Section 4.3). In
particular Section~4.3 considers a PABM setting wherein
$\lambda_{kk}$ is a vector in $\mathbb{R}^{n_k}$ whose elements
are i.i.d $\mathrm{Beta}(1,2)$ random variables and $\lambda_{k\ell}$
(for $k \not = \ell$) is a
vector in $\mathbb{R}^{n_k}$ whose elements are i.i.d $\mathrm{Beta}(2,1)$
random variables. This result in an average within-community edge
probability of $1/3$ and a between-community edge
probability of $2/3$. The results for this simultation setting are
very similar to those presented in the original manuscript and is also
consistent with the theoretical results in the paper.
}  
\textcolor{blue}{
On a related note, it is our view that the distinction between
assortative and disassortative is not always meaningful for the
PABM. Indeed, unlike the SBM, in the PABM each vertex is free to have a higher affinity to its
own community or to other communities. For example, suppose vertex $i$
belongs to community $1$. Then the average probability of an edge
between vertex $i$ and another vertex in community $1$ is $n^{-1}
\rho_n 
\lambda_{i1} \sum_{j} \lambda_{j1}$ while the average probability of an edge between vertex $i$ and another vertex in community $k \not =
1$ is $n^{-1} \rho_n \lambda_{ik} \sum_{j} \lambda_{j1}$ and thus, on
average, vertex $i$ have more affinity with vertices in community $k \not =
1$ whenever $\lambda_{i1} < \lambda_{ik}$
}

@@

17. Sections 4 and 5. Why do you use either the ARI criterion or the misclassification rate to assess the recovery of the nodes clustering?  
\textcolor{blue}{
The first two datasets that we analyzed in section 5 had also been analyzed
in previous papers on the PABM. In particular the
Leeds Butterfly dataset was analyzed in \citep{noroozi2019estimation}
and while \citep{noroozi2019estimation} provides enough details
for us to preprocess the data to match their analysis, they do not
provide code to run their clustering algorithm (which is based on SSC
applied to the rows of $A$). Therefore we used their reported
benchmark values, which are given in terms of ARI. 
In order to make an apples-to-apples comparison, we also report the ARI for OSC and SSC-ASE here as well. 
For all other analyses (simulation and real data), we use the
misclassification error rate/count. 
}

18. Do you consider sparsity in the simulated adjacency matrices?  
\textcolor{blue}{
We have added another simulation in the supplemental materials. 
In this simulation, we fix $n = 2048$ and $K = 3$ and vary $\rho \in
(0, 1)$. In particular we see that the performance of all algorithms
deteriorate significantly as the graphs become sparser. Nevertheless
it appears that the accuracy of OSC is always comparable and/or better
than the remaining algorithms for all levels of the sparsity
considered.
}

19. Why do you limit your simulation to assortative structure? 
Is it possible to see the results with a larger value of $K$? 
To what extent does the value of $K$ impact the computational burden for the algorithms?  
\textcolor{blue}{
As we discussed in the response to comment
$14$, we have added a third simulation to address the issue of disassortativity (see Section
4.3 of the revised manuscript). The performance of all algorithms
for the setting in Section~4.3 is qualitatively similar to those
presented earlier in Section~4.1 and Section~4.2 and is consistent with the theory. 
}  
\textcolor{blue}{
As for the computational complexity of the algorithms, the main bottleneck for OSC and
SSC-ASE is the construction of the affinity matrix $B$. 
For OSC, this involves a (truncated) spectral decomposition to extract the
$K^2$ largest eigenvalues in modulus (and their corresponding
eigenvectors) followed by a matrix multiplication of $n \times K^2$
and $K^2 \times n$ matrices, so the time complexity is of the order $O(n^2 K^2)$. 
For SSC-ASE (applying SSC on the rows of the adjacency spectral
embedding $\hat{V}$) we again start with spectral decomposition, followed by
$n$ LASSO problems with a design matrix of size 
$K^2 \times (n - 1)$, so the complexity is $O(n^2 K^2 + n^2 K^4 + n K^6)$ \cite{10.1214/009053604000000067}. 
SSC-A (applying SSC on the rows of $A$) as proposed in \cite{noroozi2019estimation} involves solving $n$ LASSO regression problems each with design matrices of size $n \times (n - 1)$, so the complexity here is $O(n^4)$. 
In practice the choice of parameter $K$ does not affect
runtimes too much. For the simulation studies we only considered $K
\leq 4$ as values of $K \geq 5$ will in general, require larger values of $n$ to achieve 
comparable estimation accuracy as $K \leq 4$.  Indeed, consider for
example estimating the entries of the edge probabilities matrix
$P$. Then the rank of $P$ is equal to $K^2$ and hence, for a given $n$, estimation of
$P$ when $K = 4$ is considerably simpler than estimation of $P$ when
$K = 6$. Cursory numerical experiments suggest that OSC and
SSC-ASE behave similarly for larger $K$ provided that $n$ is large enough. 
}

@@

The second part of the paper investigates the implications of this relation to the GRDPG (and the identified latent positions) on estimation algorithms for the PABM. 
New algorithms for community detection and parameter estimation are proposed. 
In particular, a new community detection method, named orthogonal spectral clustering (OSC), is proposed that directly exploits the result of Theorem 1. 
Furthermore, an improvement of the sparse subspace clustering (SSC) algorithm is proposed consisting of applying SSC to the adjacency spectral embedding of the network (ASE), referred to as SSC-ASE. For both algorithms it is shown that the subspace detection property holds and that the convergence rate of the detection error is faster than for classical SSC. 
Likewise, it is shown that parameter estimation can be improved. 
A numerical study illustrates the performance of the algorithms in comparison with existing methods. 
It shows that OSC performs the best, and also that SSC-ASE improves the results with respect to classical SSC in some scenarios. 
In the application on real datasets, OSC and SSC-ASE provide good results, but do not achieve the performance of modularity maximization (MM).

The topic of the paper is interesting as it concerns a recent graph model and the results are valuable. 
Contributions are two-fold: theoretic and computational. 
The paper is well written, and the presentation of the results and the proofs is very clear.

I only have some minor comments and a couple of questions:

### 2022-02-08 referee-response.Rmd

* I have a general issue with latent position models like the GRDPG due to the nonidentifiability of the latent space. 
As stated in the paper, latent positions are not unique, and not even the dimension of the latent space is identifiable. 
Theorem 1 (or its proof) provide one of all possible latent positions. 
A general question is what is the consequence of this specific choice. 
Does the latent space (or its dimension) have an impact on the clustering results? 
Indeed, I lack intuition about what type of latent space would be optimal for community detection or estimation in the PABM. 
The space with the smallest possible dimension? 
Do the authors have an idea of the properties of the latent space that would give the best results for community detection? 
Related to this issue, is the latent space of the proof of Theorem 1 the smallest possible latent space to relate PABM to the GRDPG?  
\textcolor{blue}{
We thank the referee for the questions. 
While it is true that there are an infinite number of latent configurations that generate the PABM, based on Theorem 1 and GRDPG results by \citet{rubindelanchy2017statistical}, we can conclusively say the following: 
\\1. If $P$ is the edge probabilities matrix for a PABM with $K$
blocks then $P$ is of rank $K^2$ and furthermore $P$ has $K (K + 1) /
2$ positive eigenvalues and $K (K - 1) / 2$ negative eigenvalues. 
\\2. Therefore, viewed as a GRDPG, the latent vectors $X$ can be
represented as vectors in $\mathbb{R}^{K^2}$ and 
$K^2$ is the smallest possible dimension. Furthermore the latent
vectors will always lie on the union of $K$ disjoint $K$-dimensional subspaces---in the configuration outlined by Theorem 1, these subspaces are orthogonal, but the multiplication by arbitrary $Q \in \mathbb{O}(p, q)$ that results in the same $P$ may skew the subspaces making them no longer orthogonal (although they will still be $K$ $K$-dimensional subspaces), 
\\3. The matrix $B = n V V^\top$, where $V$ is the $n \times K^2$
matrix of eigenvectors of $P$ (corresponding to the non-zero
eigenvalues of $P$), is such that $B_{ij} = 0$ if and only
if vertices $i$ and $j$ are in different communities, i.e.,
$v_i^{\top} v_j = 0$ if and only if vertices $i$ and $j$ are in
different communities; here $v_i$ denote the $i$th row of $V$. 
}  
\textcolor{blue}{
Therefore, the dimensionality of the latent configuration is not in question, only what kind of effect multiplication by unidentifiable $Q$ has on the configuration. 
The OSC algorithm circumvents the nonidentifiability of the latent
configuration by only considering the entries of $\hat{B} = n \hat{V}
\hat{V}^{\top}$ as $n\hat{V}\hat{V}^{\top}$ is, by the asymptotic properties of
adjacency spectral embedding, close to $n VV^{\top}$ and $n V
V^{\top}$ is always unique (the orthogonal projection onto a column
space is always unique). Similarly, SSC-ASE also circumvents the
non-identifiability of the latent positions
by using only the rows of $n^{1/2} \hat{V}$ as two rows $\hat{v}_i$ and
$\hat{v}_j$ belonging to different communities will 
have $n \hat{v}_i^{\top} \hat{v}_j \approx 0$. 
}  
\textcolor{blue}{
That said, if we look at the latent configuration in Theorem 1 then we
see that a $K$-dimensional (not $K^2$-dimensional) embedding is
sufficient for clustering, i.e., the embedding corresponding to the
diagonal blocks $\lambda_{11}, \lambda_{22}, \dots, \lambda_{KK}$. 
If we are able to find this embedding then this would greatly reduce
the number of dimensions required for the community
recovery/clustering. In practice, this isn't always possible due to
the nonidentifiability issue; in particular, determining which of the
$K$ out of $K^2$ eigenvectors should be used is the main difficulty.
Nevertheless we have found via numerical experiments that accuracte community
recovery is generally possible when we use only the eigenvectors corresponding
to the $2^\mathrm{nd}$ to the $(K+1)^\mathrm{st}$ largest eigenvalues
of $A$. 
}

### 2022-02-08 referee-response.Rmd

10. p8, line 14-21. I don't see why the permutation given by $\Pi$ has $K$ fixed points.  
\textcolor{blue}{
Example 1 illustrates these fixed points explicitly for $K = 3$. 
More specifically, comparing the matrices $X$ and $Y$, we see that
they have three columns in common: the first, fifth, and ninth. 
The other columns are permuted. The proof of Theorem 1 then
generalizes this to arbitrary $K$. In particular by the construction
of $X$, each column of $X$ has exactly one non-zero block
corresponding to some vector $\lambda_{k \ell} \in \mathbb{R}^{n_k}$. The fixed
points of $\Pi$ are then the columns with indices $r(K+1) +1 $ for $0 \leq r
\leq K-1$; these columns contain the vectors $\lambda_{kk}$ for
$1 \leq k \leq K$. The cycles of order $2$ swap the columns of $X$
containing the vectors $\lambda_{k \ell}$ and $\lambda_{\ell k}$ for
the $\tbinom{K}{2}$ pairs $\{k,\ell\}$ with $k \not = \ell$. 
}

11. Section 3.1. I found it confusing that community detection algorithms are detailed when $P$ is assumed to be known and then methods for estimating $P$ are recalled. A sentence in the beginning of this section to expose what is presented could help the reader.  
\textcolor{blue}{
We thank the referee for pointing out this issue. We had revised the
manuscript to include the line "In our methods, the data that are
observed is only the adjacency matrix $A \sim
\mathrm{PABM}(\{\lambda^{(k \ell)}\}_K, \rho_n)$ along with an assumed
number of communities, $K$. To motivate our methods, we first
consider community detection and parameter estimation in the case
where we know the edge probability matrix $P$ beforehand, noting
that community memberships and popularity parameters are not
immediately discernible from $P$ itself. After establishing methods
for community detection and parameter estimation from $P$, we use the
consistency property of ASE to demonstrate that the same methods work
for $A$ almost surely as $n \to \infty$." Indeed, the assumption of
known $P$ helps motivate Algorithm $1$ through Algorithm~3 and the
theoretical results in Theorem~2. Theorem 3 and Theorem 4 then follows
by leveraging the consistency properties for adjacency spectral embedding
to show that community recovery and parameters estimation using $A$ is asymptotically equivalent to
community recover and parameters estimation using the true but unknown $P$. 
}

@@

14. p11, line 31. "... SSC on the spectral embedding of $A$". Could you support this assertion by references?  
\textcolor{blue}{
There are no references for this, as it is, to the best of our knowledge, a novel idea. 
We emphasize that while SSC has been proposed for community detection
for PABM (see e.g., \citep{noroozi2019estimation}), these work apply SSC
directly on the rows of $A$ (which are binary vectors in
$\mathbb{R}^{n}$) rather than on the ASE of $A$ (which are real-valued
vectors in $\mathbb{R}^{d}$ where $d \ll n$). The motivation behind
our use of SSC comes  from Theorem 1 wherein we show that the latent
vectors for PABM lie on the union of orthogonal subspaces in $\mathbb{R}^{d}$ and hence,
by the consistency properties of ASE, the rows of the estimated
eigenvectors $\hat{V}$ also lies close to the union of orthogonal
subspaces in $\mathbb{R}^{d}$. In contrast, while the rows of $P$ do lie exactly on the
union of $K$ subspaces in $\mathbb{R}^{n}$, the rows of $A$
(which are now noisy binary vectors) could be quite far from the
corresponding rows of $P$ and hence need not lie close to the union
of any $K$ subspaces in $\mathbb{R}^{n}$ 
}

15. p12, line 13. It could help recall what the $z_i, z_j$ are.  
\textcolor{blue}{
We have changed $z_i$ to ``community labels''.
}

16. p14. Do you think that the OSC algorithm can retrieve other kinds of structure than assortative ones?  
\textcolor{blue}{
The OSC algorithm works for any PABM parameters setting and is thus
agnostic to the particular structure (such as assortativity vs
disassortativity). We had revised the manuscript to include an
additional simulation demonstrating this claim (see Section 4.3). In
particular Section~4.3 considers a PABM setting wherein
$\lambda_{kk}$ is a vector in $\mathbb{R}^{n_k}$ whose elements
are i.i.d $\mathrm{Beta}(1,2)$ random variables and $\lambda_{k\ell}$
(for $k \not = \ell$) is a
vector in $\mathbb{R}^{n_k}$ whose elements are i.i.d $\mathrm{Beta}(2,1)$
random variables. This result in an average within-community edge
probability of $1/3$ and a between-community edge
probability of $2/3$. The results for this simulation setting are
very similar to those presented in the original manuscript and is also
consistent with the theoretical results in the paper.
}  
\textcolor{blue}{
On a related note, it is our view that the distinction between
assortative and disassortative is not always meaningful for the
PABM. Indeed, unlike the SBM, in the PABM each vertex is free to have a higher affinity to its
own community or to other communities. For example, suppose vertex $i$
belongs to community $1$. Then the average probability of an edge
between vertex $i$ and another vertex in community $1$ is $n^{-1}
\rho_n 
\lambda_{i1} \sum_{j} \lambda_{j1}$ while the average probability of an edge between vertex $i$ and another vertex in community $k \not =
1$ is $n^{-1} \rho_n \lambda_{ik} \sum_{j} \lambda_{j1}$ and thus, on
average, vertex $i$ have more affinity with vertices in community $k \not =
1$ whenever $\lambda_{i1} < \lambda_{ik}$
}

@@

The second part of the paper investigates the implications of this relation to the GRDPG (and the identified latent positions) on estimation algorithms for the PABM. 
New algorithms for community detection and parameter estimation are proposed. 
In particular, a new community detection method, named orthogonal spectral clustering (OSC), is proposed that directly exploits the result of Theorem 1. 
Furthermore, an improvement of the sparse subspace clustering (SSC) algorithm is proposed consisting of applying SSC to the adjacency spectral embedding of the network (ASE), referred to as SSC-ASE. For both algorithms it is shown that the subspace detection property holds and that the convergence rate of the detection error is faster than for classical SSC. 
Likewise, it is shown that parameter estimation can be improved. 
A numerical study illustrates the performance of the algorithms in comparison with existing methods. 
It shows that OSC performs the best, and also that SSC-ASE improves the results with respect to classical SSC in some scenarios. 
In the application on real datasets, OSC and SSC-ASE provide good results, but do not achieve the performance of modularity maximization (MM).

The topic of the paper is interesting as it concerns a recent graph model and the results are valuable. 
Contributions are two-fold: theoretic and computational. 
The paper is well written, and the presentation of the results and the proofs is very clear.

I only have some minor comments and a couple of questions:

* I have a general issue with latent position models like the GRDPG due to the nonidentifiability of the latent space. 
As stated in the paper, latent positions are not unique, and not even the dimension of the latent space is identifiable. 
Theorem 1 (or its proof) provide one of all possible latent positions. 
A general question is what is the consequence of this specific choice. 
Does the latent space (or its dimension) have an impact on the clustering results? 
Indeed, I lack intuition about what type of latent space would be optimal for community detection or estimation in the PABM. 
The space with the smallest possible dimension? 
Do the authors have an idea of the properties of the latent space that would give the best results for community detection? 
Related to this issue, is the latent space of the proof of Theorem 1 the smallest possible latent space to relate PABM to the GRDPG?  
\textcolor{blue}{
We thank the referee for the questions. 
While it is true that there are an infinite number of latent configurations that generate the PABM, based on Theorem 1 and GRDPG results by \citet{rubindelanchy2017statistical}, we can conclusively say the following: 
\\1. If $P$ is the edge probabilities matrix for a PABM with $K$
blocks then $P$ is of rank $K^2$ and furthermore $P$ has $K (K + 1) /
2$ positive eigenvalues and $K (K - 1) / 2$ negative eigenvalues. 
\\2. Therefore, viewed as a GRDPG, the latent vectors $X$ can be
represented as vectors in $\mathbb{R}^{K^2}$ and 
$K^2$ is the smallest possible dimension. Furthermore the latent
vectors will always lie on the union of $K$ disjoint $K$-dimensional subspaces---in the configuration outlined by Theorem 1, these subspaces are orthogonal, but the multiplication by arbitrary $Q \in \mathbb{O}(p, q)$ that results in the same $P$ may skew the subspaces making them no longer orthogonal (although they will still be $K$ $K$-dimensional subspaces), 
\\3. The matrix $B = n V V^\top$, where $V$ is the $n \times K^2$
matrix of eigenvectors of $P$ (corresponding to the non-zero
eigenvalues of $P$), is such that $B_{ij} = 0$ if and only
if vertices $i$ and $j$ are in different communities, i.e.,
$v_i^{\top} v_j = 0$ if and only if vertices $i$ and $j$ are in
different communities; here $v_i$ denote the $i$th row of $V$. 
}  
\textcolor{blue}{
Therefore, the dimensionality of the latent configuration is not in question, only what kind of effect multiplication by unidentifiable $Q$ has on the configuration. 
The OSC algorithm circumvents the nonidentifiability of the latent
configuration by only considering the entries of $\hat{B} = n \hat{V}
\hat{V}^{\top}$ as $n\hat{V}\hat{V}^{\top}$ is, by the asymptotic properties of
adjacency spectral embedding, close to $n VV^{\top}$ and $n V
V^{\top}$ is always unique (the orthogonal projection onto a column
space is always unique). Similarly, SSC-ASE also circumvents the
non-identifiability of the latent positions
by using only the rows of $n^{1/2} \hat{V}$ as two rows $\hat{v}_i$ and
$\hat{v}_j$ belonging to different communities will 
have $n \hat{v}_i^{\top} \hat{v}_j \approx 0$. 
}  
\textcolor{blue}{
That said, if we look at the latent configuration in Theorem 1 then we
see that a $K$-dimensional (not $K^2$-dimensional) embedding is
sufficient for clustering, i.e., the embedding corresponding to the
diagonal blocks $\lambda_{11}, \lambda_{22}, \dots, \lambda_{KK}$. 
If we are able to find this embedding then this would greatly reduce
the number of dimensions required for the community
recovery/clustering. In practice, this isn't always possible due to
the nonidentifiability issue; in particular, determining which of the
$K$ out of $K^2$ eigenvectors should be used is the main difficulty.
Nevertheless we have found via numerical experiments that accurate community
recovery is generally possible when we use only the eigenvectors corresponding
to the $2^\mathrm{nd}$ to the $(K+1)^\mathrm{st}$ largest eigenvalues
of $A$. 
}
